---
title: "AM Project"
format: html
---

# Preamble
```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(vroom)
library(readr)
library(dplyr)
library(lubridate)
```

```{r}
i_am("AM PROJECT.Rproj")
```

# Construction of singular datasets

## CPI (2008=100)

```{r}
cpi_df_full <- vroom(
  "Data/CPI.csv",
  show_col_types = FALSE
) |>
  rename(Month = observation_date, CPI = CPIAUCSL_NBD20080101) |>
  mutate(
    Month = as.Date(Month),
    CPI = as.numeric(CPI)
  ) |>
  arrange(Month) |>
  mutate(
    CPI = ifelse(is.na(CPI), (lag(CPI) + lead(CPI)) / 2, CPI)
  ) 

start_month <- min(cpi_df_full$Month)
end_month <- max(cpi_df_full$Month)

expected_months <- seq(start_month, end_month, by = "month")

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = nrow(cpi_df_full),
  n_months_expected = length(expected_months),
  coverage_ratio = nrow(cpi_df_full) / length(expected_months)
)

stats

missing_months <- setdiff(expected_months, cpi_df_full$Month)

missing_months

ggplot(cpi_df_full, aes(x = Month, y = CPI)) +
  geom_line() +
  theme_minimal()
```


## Production Dataset

The production dataset is sourced from the U.S. Energy Information Administration (EIA) and reports world crude oil production including lease condensate, measured in million barrels per day. The data are provided at monthly frequency and represent aggregate global production levels as compiled by the EIA. Source: https://www.eia.gov/international/data/world

```{r}
prod_df_full <- vroom(
  "Data/Crude_Oil_Including_Lease_Condensate_Production.csv",
  col_names = FALSE,
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
)

prod_df_full <- tibble(
  Month = as.Date(gsub("^X", "", unlist(prod_df_full[1, ])), format = "%Y-%m-%d"),
  Production = as.numeric(gsub(",", ".", unlist(prod_df_full[2, ])))
) |>
  arrange(Month) |>
  mutate(
    dprod = (log(Production) - lag(log(Production)))
  ) |>
  select(Month, dprod) |>
  drop_na()

start_month <- min(prod_df_full$Month)
end_month <- max(prod_df_full$Month)

expected_months <- seq(start_month, end_month, by = "month")

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = nrow(prod_df_full),
  n_months_expected = length(expected_months),
  coverage_ratio = nrow(prod_df_full) / length(expected_months)
)

stats

missing_months <- setdiff(expected_months, prod_df_full$Month)

missing_months

ggplot(prod_df_full, aes(x = Month, y = dprod)) +
  geom_line() +
  theme_minimal()

```

## Real Economic Activity

The real economic activity (REA) measure is obtained from the Federal Reserve Bank of Dallas. This index is the global real economic activity indicator originally proposed by Kilian and is constructed using dry bulk shipping freight rates to capture fluctuations in global demand for industrial commodities. The series used in this analysis corresponds to the updated version released in 2019, which incorporates methodological refinements and data revisions discussed in Kilian (2019). Sources: https://www.dallasfed.org/research/igrea ; https://www.sciencedirect.com/science/article/abs/pii/S0165176519300746?via%3Dihub


```{r}
rea_df_full <- vroom(
  "Data/Real_Economic_Activity_Index.csv",
  locale = locale(decimal_mark = ","),
  show_col_types = FALSE
) |>
  rename(Month = 1, rea = 2) |>
  mutate(
    Month = as.Date(Month),
    rea = as.numeric(rea)
  ) |>
  select(Month, rea)|>drop_na()

start_month <- min(rea_df_full$Month)
end_month   <- max(rea_df_full$Month)

n_obs <- nrow(rea_df_full)
n_expected <- length(seq(start_month, end_month, by = "month"))
coverage <- n_obs / n_expected

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = n_obs,
  n_months_expected = n_expected,
  coverage_ratio = coverage
)

stats

ggplot(rea_df_full, aes(x = Month, y = rea)) +
  geom_line() +
  theme_minimal()
```

## (Real) Price of Crude Oil

The real price of crude oil is constructed using the refiner acquisition cost of imported crude oil, as reported by the U.S. Energy Information Administration (EIA). This nominal price series is deflated using the U.S. Consumer Price Index (CPI, 2008 = 100) to obtain a measure of the real price of crude oil. The resulting series reflects inflation-adjusted movements in crude oil prices faced by U.S. refiners. Sources: https://www.eia.gov/dnav/pet/pet_pri_rac2_dcu_nus_m.htm ; https://fred.stlouisfed.org/series/CPIAUCSL
```{r}
price_df_full <- vroom(
  "Data/U.S._Crude_Oil_Imported_Acquisition_Cost_by_Refiners.csv",
  delim = ";",
  col_names = FALSE,
  locale = locale(decimal_mark = ","),
  show_col_types = FALSE
)|>
  rename(
    Month = X1,
    Price = X2
  ) |>
  mutate(
    Month = as.Date(Month),
    Price = as.numeric(Price)
  ) |>arrange(Month)|>
  left_join(cpi_df_full, by = "Month") |>
  mutate(
    rpo =log((Price / CPI))
  ) |>
  select(Month, rpo) |>
  drop_na()

start_month <- min(price_df_full$Month)
end_month   <- max(price_df_full$Month)

n_obs <- nrow(price_df_full)
n_expected <- length(seq(start_month, end_month, by = "month"))
coverage <- n_obs / n_expected

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = n_obs,
  n_months_expected = n_expected,
  coverage_ratio = coverage
)

stats

ggplot(price_df_full, aes(x = Month, y = rpo)) +
  geom_line() +
  theme_minimal()
```

## (Real) WTI Price

```{r}
wti_df_full <- vroom(
  "Data/WTI_daily.csv",
  locale = locale(decimal_mark = ","),
  show_col_types = FALSE
) |>
  rename(Month = month)

wti_df_full <- wti_df_full |>
  mutate(Month = as.Date(Month)) |>
  mutate(ym = format(Month, "%Y-%m")) |>
  group_by(ym) |>
  filter(Month == max(Month)) |>
  ungroup() |>
  mutate(Month = as.Date(paste0(ym, "-01"))) |>
  select(-ym)

wti_df_full <- wti_df_full |>
  left_join(cpi_df_full, by = "Month") |>
  mutate(
    wti_real = log(WTI_day / CPI)
  )|>select(Month, wti_real)

wti_df_full <- wti_df_full |>
  arrange(Month) |>
  mutate(
    wti_real = if_else(
      is.na(wti_real),
      (lag(wti_real) + lead(wti_real)) / 2,
      wti_real
    )
  )

start_month <- min(wti_df_full$Month)
end_month   <- max(wti_df_full$Month)

n_obs <- nrow(wti_df_full)
n_expected <- length(seq(start_month, end_month, by = "month"))
coverage <- n_obs / n_expected

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = n_obs,
  n_months_expected = n_expected,
  coverage_ratio = coverage
)

stats

ggplot(wti_df_full, aes(x = Month, y = wti_real)) +
  geom_line() +
  theme_minimal()
```
## GDP and CPI

```{r}
gdp_real_df_full <- vroom(
  "Data/GDP_Real.csv",
  col_names = FALSE,
  delim = ";",
  locale = locale(decimal_mark = ",")
)|>
  rename(
    Month = X1,
    gdp_real = X2
  ) |>
  mutate(
    Month = as.Date(Month),
    gdp_real = as.numeric(gdp_real)
  )

gdp_real <- gdp_real_df_full |>
  filter(Month >= as.Date("1974-01-01"))

write.table(
  gdp_real,
  file = "MATLAB/Replication_Extended/gdp_extended.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

gdp_real <- gdp_real_df_full |>
  filter(
    Month >= as.Date("1974-01-01"),
    Month < as.Date("2008-01-01")
  )

write.table(
  gdp_real,
  file = "MATLAB/Replication/gdp_replication.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

cpi <- cpi_df_full |>
  filter(Month >= as.Date("1974-01-01"))

write.table(
  cpi,
  file = "MATLAB/Replication_Extended/cpi_extended.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

cpi <- cpi_df_full |>
  filter(
    Month >= as.Date("1974-01-01"),
    Month < as.Date("2008-01-01")
  )

write.table(
  cpi,
  file = "MATLAB/Replication/cpi_replication.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)


```
# Merging Datasets

```{r}
base_dataset_replication <- prod_df_full |>
  inner_join(rea_df_full, by = "Month") |>
  inner_join(price_df_full, by = "Month") |>
  arrange(Month)

start_month <- min(base_dataset_replication$Month)
end_month   <- max(base_dataset_replication$Month)

n_obs <- nrow(base_dataset_replication)
n_expected <- length(seq(start_month, end_month, by = "month"))
coverage <- n_obs / n_expected

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = n_obs,
  n_months_expected = n_expected,
  coverage_ratio = coverage
)

stats
```

## Replication Datasets
```{r}

# Extended Dataset (1974.1-2025.9)

kilian_replication <- base_dataset_replication |>
  select(dprod, rea, rpo)

write.table(
  kilian_replication,
  file = "MATLAB/Replication_Extended/kilian_replication_extended.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

# Original Dataset (1974.1-2007.12)

kilian_replication <- base_dataset_replication |>
  filter(Month < as.Date("2008-01-01")) |>
  select(dprod, rea, rpo)

write.table(
  kilian_replication,
  file = "MATLAB/Replication/kilian_replication.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

# New Dataset (2006.1-2025.9)

kilian_replication_2006_2025 <- base_dataset_replication |>
  filter(Month >= as.Date("2006-01-01")) |>
  select(dprod, rea, rpo)

write.table(
  kilian_replication_2006_2025,
  file = "MATLAB/Replication_2006_2025/kilian_replication_2006_2025.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

```

### Kilian Original Datset

```{r}
data_kilian_original <- read_table(
  "Data/Original_Kilian_Dataset.txt",
  col_names = FALSE,
  col_types = cols(.default = col_double())
)

n_obs <- nrow(data_kilian_original)

start_date <- as.Date("1973-02-01")
date_seq <- seq(from = start_date, by = "month", length.out = n_obs)

data_kilian_original <- data_kilian_original |>
  mutate(Month = date_seq) |>
  select(Month, everything()) |>
  rename(
    dProd = X1,
    REA   = X2,
    RPO   = X3
  )

# sanity check (replication check)
data_kilian_original |>
  summarise(
    mean_dProd = mean(dProd),
    mean_REA   = mean(REA),
    mean_RPO   = mean(RPO)
  )

compare <- data_kilian_original |>
  left_join(base_dataset_replication, by = "Month")

# dProd
compare |>
  ggplot(aes(x = Month)) +
  geom_line(aes(y = dProd), color = "green") +
  geom_line(aes(y = dprod))

# REA
compare |>
  ggplot(aes(x = Month)) +
  geom_line(aes(y = REA), color = "green") +
  geom_line(aes(y = rea))

# RPO
compare |>
  ggplot(aes(x = Month)) +
  geom_line(aes(y = RPO), color = "green") +
  geom_line(aes(y = rpo))

compare_to_export <- compare |>
  select(Month, dProd, REA, RPO, dprod, rea, rpo)

write.table(
  compare_to_export,
  file = "MATLAB/Data/compare.txt",
  sep = " ",
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE
)

```


# Extension Dataset

```{r}
expectation_df_full <- vroom(
  "Data/Expected_Crude_Oil_Inventories_US.csv",
  locale = locale(decimal_mark = ","),
  show_col_types = FALSE
) |>
  rename(Month = month) |>
  mutate(
    Month = as.Date(Month),
    across(-Month, ~ log(.))
  ) |>
  arrange(Month)

start_month <- min(expectation_df_full$Month)
end_month   <- max(expectation_df_full$Month)

n_obs <- nrow(expectation_df_full)
n_expected <- length(seq(start_month, end_month, by = "month"))
coverage <- n_obs / n_expected

stats <- tibble(
  start_month = start_month,
  end_month = end_month,
  n_months_observed = n_obs,
  n_months_expected = n_expected,
  coverage_ratio = coverage
)

stats

```

## Future Inventories Dataset

```{r}
svar4_df_full <- base_dataset_replication |>
  left_join(expectation_df_full, by = "Month") |>
  arrange(Month) |>
  filter(Month >= as.Date("2006-01-01"))

ggplot(svar4_df_full, aes(x = Month, y = `6`)) +
  geom_line()

vars_X <- as.character(1:12)

for (x in vars_X) {
  write.table(
    svar4_df_full |>
      select(dprod, rea,all_of(x), rpo) |>
      drop_na(),
    file = file.path(
      "MATLAB/Extension_Level",
      paste0("data_expected_inventory_", x, ".txt")
    ),
    row.names = FALSE,
    col.names = FALSE
  )
}
```

## Shock Proxy
```{r}
calculate_inventory_shocks <- function(data) {
  
  data <- data |> arrange(Month)
  
  shock_df <- tibble(Month = data$Month)
  
  for (h in 2:11) {
    
    col_current <- as.character(h)
    col_prev    <- as.character(h + 1)
    
    if (!(col_current %in% names(data)) || !(col_prev %in% names(data))) next
    
    shock_df[[paste0("Shock_Horizon_", h)]] <-
      data[[col_current]] - lag(data[[col_prev]], 1)
  }
  
  shock_df |> drop_na()
}

inventory_shocks_final <- calculate_inventory_shocks(expectation_df_full)

```

```{r}
library(tidyverse)

inventory_shocks_final |>
  pivot_longer(
    cols = starts_with("Shock_Horizon"),
    names_to = "Horizon",
    values_to = "Shock_Value"
  ) |>
  mutate(
    Horizon = as.numeric(str_extract(Horizon, "\\d+")),
    Horizon = factor(Horizon, levels = sort(unique(Horizon)))
  ) |>
  ggplot(aes(x = Month, y = Shock_Value, color = Horizon, group = Horizon)) +
  geom_line(linewidth = 0.6, alpha = 0.7) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) +
  labs(
    title = "Inventory Expectation Shocks Across Horizons",
    subtitle = "Horizons from 2 to 11 months",
    x = NULL,
    y = "Shock magnitude",
    color = "Horizon (months)"
  )

```

```{r}

var_datasets_list <- list()

for (h in 2:11) {
  
  shock_col_name <- paste0("Shock_Horizon_", h)
  
  dataset_name <- paste0("data", h)
  

  temp_dataset <- base_dataset_replication |>
  
    inner_join(inventory_shocks_final, by = "Month") |>
    
    
    select(dprod, rea, all_of(shock_col_name), rpo) |>
     rename(inventory_shock = all_of(shock_col_name)) |> 
    
    drop_na()
  
  var_datasets_list[[dataset_name]] <- temp_dataset
  
  assign(dataset_name, temp_dataset)
}


print(head(data2))

```


```{r}
output_path <- "MATLAB/Extension"

dir.create(output_path, recursive = TRUE, showWarnings = FALSE)

for (name in names(var_datasets_list)) {
  
  df_to_save <- var_datasets_list[[name]]
  
  filename <- file.path(output_path, paste0(name, ".txt"))
  
  write.table(
    df_to_save,
    file = filename,
    sep = " ",
    row.names = FALSE,
    col.names = FALSE,
    quote = FALSE
  )
  
  message("Salvato: ", filename)
}

```


